Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-24 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
small commit to merge

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow us to set upper limits on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during the post-merger stage. So far, none of these null tests have reported any deviation from GR.

@@

\ab{We will treat the modified theories of gravity as EFTs, and focus on 
finite-size effects (see, e.g., Ref.~\cite{Sennett:2019bpc}). Thus, for each gravity theory the fundamental lenght-scale 
$\ell \lesssim GM/c^2$, where $M$ is the mass of the BH. This implies 
that observable deviations from GR present in those theories arise from modifications 
to the Kerr geometry of each individual BH. Those finite-size effects can manifest themselves in 
the QNMs of the remnant produced by the merger, but also in the GW phasing of the inspiral through 
corrections to the GR spin-induced quadrupole and Love numbers. However, here we will not consider the latter, 
instead, we will only study the impact of finite-size effects on the QNMs of the 
remnant.}

@@

The theory admits as a solution, the garden-variety Schwarzschild BH of GR.
This is not the case when rotation is included and the Kerr metric is not a
solution of the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg} and
%
the leading-order modification to the GW phase enters at 2PN~\cite{Yagi:2011xp}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.~6 of
Ref.~\cite{LIGOScientific:2021sio}). \ab{So far, analyses that used only the inspiral portion of 
the BBH GW signals were not able to set meaningful bounds on such deviation at 2PN order~\cite{Nair:2019iur,Perkins:2021mhb}.}
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm dCS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated NS
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
NS GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}, using
equation-of-state independent relations between NS moment of inertia and tidal deformability~\cite{Yagi:2013bca,Yagi:2013awa,Gupta:2017vsl}.

@@

\ab{For the EFT of GR with dimension-eight operators, Ref.~\cite{Sennett:2019bpc} focused on the 
orbital effects (i.e., the inspiral stage) and performed 
Bayesian model selection using the two lowest-mass BBHs events of the second-observig run of 
the LIGO-Virgo Collaboration, notably GW151226 and GW170608. They found that the 
data disfavor the appearance of new physics on distance scales around $\ell_{\rm qEFT} \sim 150$ km.}

@@

For the nonrotating QNMs we used the numerical data of Ref.~\cite{Blazquez-Salcedo:2016enn} and
generated a new quartic fit in $\gamma_{\rm EdGB} = \ell_{\rm EdCS} / M$ using numerical QNM data valid for small
values of the coupling constant [see, in particular, Eq.~(27) and Fig.~2 of Ref.~\cite{Blazquez-Salcedo:2016enn}].
%
We find,
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= M {\rm Re}(\sigma)_{\rm GR} \, \left( 1 + 0.0107 \, \gamma_{\rm EdGB} \right),
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= M {\rm Im}(\sigma)_{\rm GR} \, \left( 1 - 0.0044 \, \gamma_{\rm EdGB} \right).
\end{align}
\end{subequations}
%
The small values of the numerical prefactors of $\gamma_{\rm EdGB}^4$ are a consequence of the how weakly the QNMs
of BHs in EdGB gravity deviate from their GR counterparts, even at moderate values $\gamma_{\rm EdGB} \approx 0.3$.

>>>>

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow to set upper limit on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during the post-merger stage. So far, none of these null tests have reported any deviation from GR.

@@

@@

The theory admits as a solution, the garden-variety Schwarzschild BH of GR.
This is not the case when rotation is included and the Kerr metric is not a
solution of the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg} and
%
the leading-order modification to the GW phase enters at 2PN~\cite{Yagi:2011xp}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.~6 of
Ref.~\cite{LIGOScientific:2021sio}). Thus, so far, dCS gravity has remained
unconstrained by works which use only on the inspiral portion of BBH GW signals~\cite{Nair:2019iur,Perkins:2021mhb}.
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm dCS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated NS
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
NS GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}, using
equation-of-state independent relations between NS moment of inertia and tidal deformability~\cite{Yagi:2013bca,Yagi:2013awa,Gupta:2017vsl}.

@@

In general \dots
%
\hscomm{Alessandra, can you include a short description of~\cite{Sennett:2019bpc} for the quartic EFT?}

@@

For the nonrotating QNMs we used the numerical data of Ref.~\cite{Blazquez-Salcedo:2016enn} and
generated a new quartic fit in $\gamma_{\rm EdGB} = \ell_{\rm EdCS} / M$ using numerical QNM data valid for small
values of the coupling constant [see, in particular, Eq.~(27) and Fig.~2 of Ref.~\cite{Blazquez-Salcedo:2016enn}].
%
We find,
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= M {\rm Re}(\sigma)_{\rm GR} \, \left( 1 + 0.0107 \, \gamma_{\rm EdGB} \right),
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= M {\rm Im}(\sigma)_{\rm GR} \, \left( 1 - 0.0044 \, \gamma_{\rm EdGB} \right).
\end{align}
\end{subequations}
%
The small values of the numerical prefactors of $\gamma_{\rm EdGB}$ are a consequence of the how weakly the QNMs
of BHs in EdGB gravity deviate from their GR counterparts, even at moderate values $\gamma_{\rm EdGB} \approx 0.3$.
